---
title: "Exploring Shiny"
author: "Robert Olendorf"
date: "March 2, 2016"
output: html_document
---

```{r. setup, echo=FALSE, warning=FALSE, message=FALSE}
  if(!require("pander"))
  {
    install.packages("pander", repos="http://cran.rstudio.com/", dependencies = TRUE)
  }
  
  library("pander")
```

## Getting Started

Creating a new Shiny App is very easy. Just go to **[File/New Project .../New Directory/Shiny App]**. Then give it a name **_random_thoughts_** this time. Go ahead and initialize the Git repository too. RStudio will create two files for you, **_ui.R** and **_server.R_**. 

### Shiny UI

The Shiny UI can be thought of consisting of a layout, inputs and outputs. We don't have enough time to go through all the options, but there is excellent documentation following the links below. You can also go to the very handy [Shiny Reference Page](http://shiny.rstudio.com/reference/shiny/latest/)

#### Layouts


  * http://shiny.rstudio.com/tutorial/lesson2/
  * http://shiny.rstudio.com/articles/layout-guide.html
  
```{r, layouts-table, echo=FALSE}
  layouts.data <- read.csv("assets/data/layouts.csv")
  pander(layouts.data)
```

#### Widgets (inputs): 

  * http://shiny.rstudio.com/tutorial/lesson3/
    
```{r, widgets-table, echo=FALSE}

  widget.data <- read.csv("assets/data/widgets.csv")
  pander(widget.data)
```
    
### Outputs:
  * http://shiny.rstudio.com/tutorial/lesson4/
  
```{r, outputs-table, echo=FALSE}
  outputs.data <- read.csv("assets/data/outputs.csv")
  pander(outputs.data)
```
    
The layouts, as you would expect, determine where everythign goes. Inputs, or Widgets, are elements that allow user input. When an input changes, the data is sent to the **_server.R_** script. **_server.R_** reanalyzes the data and sends the results to the correct output function.

In the default application the RStudio creates, we have a ```fluidPage``` with a ```titlePanel``` using a  ```sidebarLayout``` with one ```sidebarPanel``` and one ```mainPanel``` defined. We have one input, a ```sliderInput()```. One ```plotOutput()```.  Inputs vary a bit in their signature, but they all require 
  
  * a **name**, used by R to identify the widget's value to the server. It is not seen by the user.
  * a **label** Seen by the user, but not used by R.
  
  Outputs rquire a *name* which identifies the data sent by the server tothe UI.
  
### Shiny Server

A server is an R script. Inputs from the UI are designated by ```input$name```, where the name is the name givin in the **Input Widget**. In the default application we can see it used by the server to define the number of bins in the ```hist()``` function.

The output is similarly defined by ```output$name```, where the name is used by the UI to display the data.

As a sanity check, lets run this application locally, then we can deploy it to ShinyApps.io too. To run the app locally just click **_Run App_** at the top center. Publishing your app is almost as easy. Click the funny circle icon next to **_Run App_**. You may have to set up your connection. If you have a Github account, its probably easiest to just user the Github login method. Then follow the instructions on connecting to ShinyApps.io.

Once you have the set up, click the publish icon. The defaults are probably fine. Shiny and RStudio will do most of the work for  you.

Once it's all working. Go back to to [ShinyApps](https://www.shinyapps.io), log in if you need to, go to your **_Dashboard_** and click on the **_Applications_** tab on the left. Choose your application, click on the **_Settings (Gear)_** icon, and you can explore a lot of the analytical tools ShinyApps.io gives you. Some of the features require you to pay. I don't, so I can't say much about them.

## Creating Our Own Application.

As part of this exercise were going to use an unnecessarily complicated dice rolling function. The function rolls a specified number of dice on a given roll and sums the values. The rolls are repeated a number of times to generate distribution of outcomes. You can also roll different sided dice! I actually didn't know there were so many kinds of dice, even *one sided!#.

I like to do things in baby steps, so lets do something similar to the default app, but add a few more widgets. Starting with the UI, lets add widgest for the number of dice, number of rolls, number of bins and number of sides.

Our UI will look somethign like this ( [v1.0 in code4lib2016_random_stochastic_chance1](https://github.com/olendorf/code4lib2016_random_stochastic_chance1/tree/v1.0)).

```r
  library(shiny)

  shinyUI(fluidPage(
    
    allowable.sides <- c(1:20, 22 , 24, 30, 32, 34, 48, 50, 60, 100, 120, 144),
  
    # Application title
    titlePanel("Random Stocastic Chance"),
  
    # Sidebar with a slider input for number of bins
    sidebarLayout(
      sidebarPanel(
        sliderInput("dice",
                    "Number of Dice",
                    min = 1,
                    max = 20,
                    value = 2),
        
        
        numericInput("rolls",
                    "Number of Rolls (Whole Numbers)",
                    min = 1,
                    value = 1000),
        
        selectInput("sides", "Number of Sides", allowable.sides),
        
        sliderInput("bins",
                    "Number of bins:",
                    min = 1,
                    max = 50,
                    value = 30),
                    
  
      # Show a plot of the generated distribution
      mainPanel(
        plotOutput("scores")
      )
    )
  ))

```

Next we need to write the server. Keeping it simple we'll just connect all the parts we defined in the UI.

```r
  


library(shiny)

dice.rolls <- function(number.dice = 1, number.rolls = 1, sides=6)
{
  output <- NULL
  if (!is.element(sides, allowable.sides))
  {
    return(paste("Error: ", toString(sides), " is not an allowable number of sides."))
  }
  
  return(replicate(number.rolls, sum(sample(1:sides, number.dice, replace = TRUE))))
  
}


allowable.sides <- c(1:20, 22 , 24, 30, 32, 34, 48, 50, 60, 100, 120, 144)

shinyServer(function(input, output) {
  
  output$scores <- renderPlot({
    scores = dice.rolls(
      number.dice = input$dice, 
      number.rolls = input$rolls,
      sides = input$sides
      )
    
    hist(scores, breaks = input$bins)
    
  })



})

```

### Improving the UI

This works pretty well but we can make some enhancements to the user's experience. First, if you play with the bins, you'll see that user defined bins don't always look so great. Lets give them the option to allow bins to be created automatically. In the UI lets add this to the code. Also lets create a grouping of UI alements, since we have some inputs that control the dice, and others the plotting. We can use the `wellPanel()` function, and add a `checkboxInput()` in `ui.R`. 

In our server we need to to refractor our code to check if the user has checked **_Bin Automatically_**. It will look something like this.

```r
  output$scores <- renderPlot({
    scores = get.scores()

    
    histogrm <- NULL
    if(input$auto_bins)
    {
      histogrm <- hist(scores)
    }
    else
    {
      histogrm <- hist(scores, breaks = input$bins)
    }
    histogrm
    
  })
```

Trying it out, it should work, but if the user choosed to **_Bin Automatically_** they can also choose to change the number of bins, but that doesn't do anything. This works as expected but is confusing, so lets disable the **_Number of bins_** slider when the **_Bin Automatically_** box is checked. Turns out this is hard in Shiny **unless** we load a new package. So load **shinyjs**. This is a nice package of javascript helpers, one of which disables inputs. To accomplish this we need to make the following changes. In `server.R` add `library(shinyjs)` near the top. Then inside the `shinyserver()` function add this.

```r
  shinyServer(function(input, output, session) {
    observe({
      if (input$auto_bins == TRUE) {
        shinyjs::disable("bins")
      } else {
        shinyjs::enable("bins")
      }
    })
  
  ...
  ...
  ...
  }
```

In `ui.R` add this jus below the `shinyUI()` method.

```r

  shinyUI(fluidPage(
    
    shinyjs::useShinyjs(),
    
    ...
    ...
    ...
  ))
```


No when we restart the server, the **_Number of bins_** input should be disabled when we check the **_Bin Automatically_** box. There is one more issue. Start increasing the **_Number of rolls_**. As you get to a million, it starts taking a long time. This makes sense, were doing a lot of work. Now if you change the number of bins, it still takes a long time. This is bad, because we're recalculating the rolls when we don't need to. We just want to rebin the data. Fortunately we can do this as well with the `reactive()` methods. We need to change our `server.R` a bit. So after all these changes it should look like below. 
You can get this on github at [tag v1.1.1]( https://github.com/olendorf/code4lib2016_random_stochastic_chance1/tree/v1.1.1).

`server.R`
```r
  library(shiny)
  library(shinyjs)
  
  dice.rolls <- function(number.dice = 1, number.rolls = 1, sides=6)
  {
    output <- NULL
    if (!is.element(sides, allowable.sides))
    {
      return(paste("Error: ", toString(sides), " is not an allowable number of sides."))
    }
    
    return(replicate(number.rolls, sum(sample(1:sides, number.dice, replace = TRUE))))
    
  }
  
  
  allowable.sides <- c(1:20, 22 , 24, 30, 32, 34, 48, 50, 60, 100, 120, 144)
  
  shinyServer(function(input, output, session) {
    
    # Observer to see if we should disable the bins slider.
    observe({
      if (input$auto_bins == TRUE) {
        shinyjs::disable("bins")
      } else {
        shinyjs::enable("bins")
      }
    })
    
    # Wrapping dice.rolls() inside the reactive method
    # and aliasing it in get.scores() provides a caching method.
    # Unless the actual inputs for dice.rolls() changes this
    # method won't get called again.
    get.scores <- reactive({
      dice.rolls(
        number.dice = input$dice, 
        number.rolls = input$rolls,
        sides = input$sides
      )
    })
    
    output$scores <- renderPlot({
      scores = get.scores()
  
      
      histogrm <- NULL
      
      # Testing if the bins should be autobinned or not.
      if(input$auto_bins)
      {
        histogrm <- hist(scores)
      }
      else
      {
        histogrm <- hist(scores, breaks = input$bins)
      }
      histogrm
      
    })
  
  
  })
```

`ui.R`
```r
  library(shiny)
  library(shinyjs)
  
  dice.rolls <- function(number.dice = 1, number.rolls = 1, sides=6)
  {
    output <- NULL
    if (!is.element(sides, allowable.sides))
    {
      return(paste("Error: ", toString(sides), " is not an allowable number of sides."))
    }
    
    return(replicate(number.rolls, sum(sample(1:sides, number.dice, replace = TRUE))))
    
  }
  
  
  allowable.sides <- c(1:20, 22 , 24, 30, 32, 34, 48, 50, 60, 100, 120, 144)
  
  shinyServer(function(input, output, session) {
    
    # Observer to see if we should disable the bins slider.
    observe({
      if (input$auto_bins == TRUE) {
        shinyjs::disable("bins")
      } else {
        shinyjs::enable("bins")
      }
    })
    
    # Wrapping dice.rolls() inside the reactive method
    # and aliasing it in get.scores() provides a caching method.
    # Unless the actual inputs for dice.rolls() changes this
    # method won't get called again.
    get.scores <- reactive({
      dice.rolls(
        number.dice = input$dice, 
        number.rolls = input$rolls,
        sides = input$sides
      )
    })
    
    output$scores <- renderPlot({
      scores = get.scores()
  
      
      histogrm <- NULL
      
      # Testing if the bins should be autobinned or not.
      if(input$auto_bins)
      {
        histogrm <- hist(scores)
      }
      else
      {
        histogrm <- hist(scores, breaks = input$bins)
      }
      histogrm
      
    })
  
  
  })

```

### Layouts

There are a lot of layouts. Just as a quick exercise, we can add tabs to our main panel. Change the main panel in `ui.R` to look like this.

```r
  mainPanel(
      tabsetPanel(type = "tabs", 
        tabPanel("Plot", plotOutput("scores")),
        tabPanel("Summary", verbatimTextOutput("summary")),
        tabPanel("Table", tableOutput("table"))
      )
      
    )
```

Then we need to change `server.R` to supply the outputs we just defined.

```r
  output$scores <- renderPlot({
      scores = get.scores()
  
      
      histogrm <- NULL
      
      # Testing if the bins should be autobinned or not.
      if(input$auto_bins)
      {
        histogrm <- hist(scores)
      }
      else
      {
        histogrm <- hist(scores, breaks = input$bins)
      }
      histogrm
      
    })
    
    output$summary <- renderPrint({summary(get.scores())})
    
    output$table <- renderTable({data.frame(scores=get.scores())})
```

Full code can be seen on Github https://github.com/olendorf/code4lib2016_random_stochastic_chance1/tree/v1.2-panels.

### Styles

We can customize even more with styles. For a quick change we can just use the **_ShinyThemes_** package. Install the package then edit `ui.R` to look like this.

```r
  library(shiny)
  library(shinythemes)
  
  shinyUI(fluidPage(theme = shinytheme("spacelab"),
  ...
  ...
  ...
  ))
```

## Making it more dynamic

Often with web applications we want it to be dynamic as much as possible. Lets try to dynamically create **Tab Panels** based on the data. We'll use the `iris` data set. Take a moment and explore it, `head(iris)`. Create a new shiny app, call it `shiny_iris`. 

Then in the `ui.R` make it look like this.

```r
  library(shiny)
  
  shinyUI(fluidPage(
  
    # Application title
    titlePanel("Old Faithful Geyser Data"),
  
    # Sidebar with a slider input for number of bins
    sidebarLayout(
      sidebarPanel(
        sliderInput("bins",
                    "Number of bins:",
                    min = 1,
                    max = 50,
                    value = 30)
      ),
  
      # Show a plot of the generated distribution
      mainPanel(
        uiOutput("irisTabs")
      )
    )
  ))
```

Notice we haven't defined our Tabs. the `uiOutput("irisTabs")` tell R to expect those UI components to be created by the Server in `output$irisTabs`. In `server.R` we need to do a lot of work. First we need to define the Tabs.

```r

 # Helper function to capitilize a word.
 capitalize <- function(str){ 
    paste(
      toupper(substring(str, 1, 1)), 
      substring(str, 2), sep="")
  }
 
 
 # Helper function to create the Tab Panels, makes using lappy a little easier.
  tabPanel.plotOutput <- function(name) {
    tabPanel(name, plotOutput(name))
  }
  
  # Gets the tabs from the data, and adds an All tab as well.
  output$irisTabs <- renderUI({
    tabs <- c("All", lapply(levels(iris$Species), capitalize))
    tabs <- lapply(tabs, tabPanel.plotOutput)
    do.call(tabsetPanel, tabs)
  })
```

If you run the application now, you should see the Tabs nicely prepared. But we don't have any data yet. This will take some work, and took me some trial and error.

```r
  # Make the "All" tab since its not in the data
  output[["All"]] <- renderPlot(
   hist(iris$Sepal.Length, main = "All Species")
  )
  
  # Get our list of the species. 
  species <- levels(iris$Species)
  
  # For eac species add the data.
  for(index in 1:length(species))
  {
   # If we don't add the local, it only shows the last plot on all the tabs.
   local({
     tab.name <- capitalize(species[index])
     subset.data <- subset(iris, Species == species[index])
     output[[tab.name]] <- renderPlot({
       hist(subset.data$Sepal.Length, main = tab.name)
     })
   })
   
   }
```


